---
title: "Getting and Cleaning Data Course Project"
output: html_notebook
---


### Review Criteria

    The submitted data set is tidy.
    The Github repo contains the required scripts.
    GitHub contains a code book that modifies and updates the available codebooks with the data to indicate all the variables and summaries calculated, along with units, and any other relevant information.
    The README that explains the analysis files is clear and understandable.
    The work submitted for this project is the work of the student who submitted it.
    
    
    
### Approach

It's important to first understand how the data is located.

- *features.txt* contains the variable names for the features. There are 561 of them. They are a the result of applying specific functions to the measurments performed to the different subjects.

- *activity_labels.txt* contains the names of the activities measured.

- */dir/subject_test.txt* Contains the column with the entries of the subjects
- */dir/X_test.txt* Contains the data for the measurements with the applied function
- */dir/Y_test.txt* Contains the column with the entries for every activity


So we take a look at the kind of files and their dimensions:


```{r}
datay <- read.table("./UCI HAR Dataset/test/y_test.txt")
dim(datay)
datax <- read.table("./UCI HAR Dataset/test/X_test.txt")
dim(datax)
datasub <- read.table("./UCI HAR Dataset/test/subject_test.txt")
dim(datasub)
features <- read.table("./UCI HAR Dataset/features.txt")
dim(features)
```

So we see that we have 2947 observations for 561 parameters.


We need to merge then

1. Features as a name row for datax and naming the datay column and the subject column
```{r}
names(datax) <- features[,2]
names(datay) <- "Activity"
names(datasub) <- "Subject"
```

2. Now we just need to get the columns with the words containing mean and standard deviation.

We can get a list of columns by using regexp:

```{r}
columns <- grep("mean()|std()", features[,2])
```

With that information we can filter now the data in datax :)

```{r}
filtered <- datax[,columns]
```

Now we just need to join the tables


