---
title: "Week 4"
output: html_notebook
---



## Editing Text Variables

### Fixing Character vectors - tolower and toupper

Let's download some example data from Baltime Public Data website.

```{r}
if(!file.exists("./data")) {dir.create("./data")}
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl,destfile="./data/cameras.csv", method="curl")
cameraData <- read.csv("./data/cameras.csv")
names(cameraData)
```

We can set all strings to lower case with 

```{r}
tolower(names(cameraData))
```

```{r}
toupper(names(cameraData))
```

### Fixing character vectors - strsplit()

```{r}
splitNames = strsplit(names(cameraData), "\\.")
splitNames[[5]]
```

```{r}
splitNames[[6]]
```

### Quick aside - lists

```{r}
mylist <- list(letters = c("A", "b", "c"), numbers = 1:3, matrix(1:25, ncol =5))
head(mylist)
```

How to access the elements?

```{r}
mylist[1]
```

This returns the first elemenet of the list, a list itself.

```{r}
mylist$letters
```

```{r}
mylist[[1]]
```


### Fixing Character vectors - sapply()

```{r}
splitNames[[6]][1]
```

```{r}
firstElement <- function(x){x[1]}
sapply(splitNames, firstElement)
```


### Peer Review Data

### Fixing character vectors - sub() / gsub()

We can use sub/gsub to substitute a character or string for another one:

```{r}
#sub/gsub("_", "", names(reviews),)
```

### Finding values - grep(), grepl()

```{r}
grep("Alameda", cameraData$intersection)
```
```{r}
table(grepl("Alameda", cameraData$intersection))
```

```{r}
cameraData2 <- cameraData[!grepl("Alameda", cameraData$intersection), ]
```

```{r}
grep("Alameda", cameraData$intersection, value=TRUE)
```


### More useful string functions

```{r}
library(stringr)
nchar("Alameda")
```

```{r}
substr("Jeffrey Leek", 1,7)
```
```{r}
paste("Jeffrey", "Leek")
```


```{r}
paste0("Jeffrey", "Leek")
```

```{r}
str_trim("Jeff    ")
```

## Regular Expressions


## Working with dates

```{r}
d1 = date()
d1
```

```{r}
d2 = Sys.Date()
d2
```

```{r}
format(d2, "%a %b %d")
```

### Creating Dates

Use as.Date to import dates that were created as text.

```{r}
weekdays(d2)
```

```{r}
months(d2)
```

```{r}
julian(d2)
```

### Lubridate

```{r}
library(lubridate); ymd("20140108")
```


```{r}
mdy("08/04/2013")
```
```{r}
dmy("03-04-2013")
```


### Dealing with times

```{r}
ymd_hms("2011-08-03 10:15:03")
```

```{r}
ymd_hms("2011-08-03 10:15:03", tz="Pacific/Auckland")
```

```{r}
?Sys.timezone
```

### Some functions have slightly different syntax

```{r}
x = dmy(c("1jan2013", "2jan2013", "31mar2013", "30jul2013"))
wday(x[1])
```

```{r}
wday(x[1], label=TRUE)
```


```{r}
?POSIXlt
```




