---
title: "Quiz 3"
output: html_notebook
---

### Question 1

The American Community Survey distributes downloadable data about United States communities. Download the 2006 microdata survey about housing for the state of Idaho using download.file() from here:

https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv

```{r}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv"
download.file(url, "./data/question1.csv")
q1data <- read.csv("./data/question1.csv")
```


and load the data into R. The code book, describing the variable names is here:

https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FPUMSDataDict06.pdf

colum ACR = 3 (>10 acres)
AGS = 6

Create a logical vector that identifies the households on greater than 10 acres who sold more than $10,000 worth of agriculture products. Assign that logical vector to the variable agricultureLogical. Apply the which() function like this to identify the rows of the data frame where the logical vector is TRUE.

```{r}
library(dplyr)
q1data <- select(q1data, ACR, AGS)
which(q1data$ACR == 3 & q1data$AGS == 6)
which(with(q1data, ACR == 3 & AGS == 6))
```


What are the first 3 values that result?


### Question 2
Using the jpeg package read in the following picture of your instructor into R

```{r}
install.packages("jpeg")
library(jpeg)
```


https://d396qusza40orc.cloudfront.net/getdata%2Fjeff.jpg

Use the parameter native=TRUE. What are the 30th and 80th quantiles of the resulting data? (some Linux systems may produce an answer 638 different for the 30th quantile)

```{r}
download.file( "https://d396qusza40orc.cloudfront.net/getdata%2Fjeff.jpg", "./data/q2.jpeg")
pic <- readJPEG("./data/q2.jpeg", native=TRUE)
quantile(pic, c(.30, .80))
```


### Question 3


```{r}
gdpdataurl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv"
edudataurl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv"

gdpdatafile <- "./data/gdp-data.csv"
edudatafile <- "./data/edu-data.csv" 

download.file(gdpdataurl, gdpdatafile)
download.file(edudataurl, edudatafile)
```


Time to import the data:

```{r}
gdp <- read.csv(gdpdatafile, header=FALSE, skip = 5, nrows = 190, blank.lines.skip = TRUE)
edu <- read.csv(edudatafile)

gdp <- select(gdp, c(1,2,4,5))
colnames(gdp) <- c("CountryCode", "ranking", "Economy", "gdp")

```

Let's now merge by CountryCode

We have two options, using join or merge


```{r}
merged <- join(gdp, edu)
merged2 <- merge(gdp, edu)
nrow(merged)
# nrow(merged2)
```

And we sort them:

```{r}
ordered <- arrange(merged, desc(ranking))
ordered2 <- arrange(merged2, desc(ranking))
ordered2$Long.Name[[13]]
```

From what we see, there seems to be 189 or 190 matches (how do we know that?) and the 13th country is St Kitts and Nevis


Result is: 189 matches, 13th country is St. Kitts and Nevis



### Question 4

What is the average GDP ranking for the "High income: OECD" and "High income: nonOECD" group?

The data is located in the Income.Group column, so we can use the *select* operations to create a new data frame with the required data.

```{r}
ranking_by_group <- select(merged, ranking, Income.Group)
oecd <- ranking_by_group %>% filter(Income.Group == "High income: OECD")
nonoecd <- ranking_by_group %>% filter(Income.Group == "High income: nonOECD")
mean(oecd$ranking)
mean(nonoecd$ranking)
```

32.96667, 91.91304


### Question 5

Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries

are Lower middle income but among the 38 nations with highest GDP?

```{r}
merged2$rankgroups <- cut(merged2$ranking, breaks=quantile(merged2$ranking))
```



Result is: 5
.